\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-26 chapters/bci.tex
Add ERP section

\fullpagefigct{figures/bci/bci_loop.pdf}{The \ac{bci} loop.
The user interacts with the \ac{bci} through a \ac{bci} paradigm, in this case involving
visual stimulation. Neural signals are recorded, and neurophysiological features
related to the paradigm are extracted. Using machine learning, a decision can
then be made based on these features, which can be presented back to the user.
In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{fig:bci-loop}

@@

The recorded brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci-loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
It would help greatly if we knew what we are looking for.
For this reason, a \ac{bci} operates using a defined \emph{paradigm}, the manner in
which interaction is performed.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
that the brain response evoked by attended stimuli is different of those
unattended,
and determine which one of a set of targets was attended.
In turn, this can be coupled to an action, like typing the letter shown on the
stimulus.
The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci-paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

@@

As a consequence, the \ac{bci} field relies heavily on the
electrical activity of the brain.
As mentioned earlier with MEG, an important concept in determining which
technology is suited for a \ac{bci} application is the \emph{invasiveness} of the technology.
As illustrated in Figure~\ref{fig:recording-modalities}, invasive electrodes can often measure a very specific brain region and can result in better signal quality, at the cost of the risks
involved with brain surgery.
\fullpagefigct{figures/bci/recording_modalities.pdf}{%
	Different electrophysiology recording modalities and \\
	their respective position in a cross-section of the skull. \\
	There is a trade-off between invasiveness and  \\
	spatial resolution.
}{fig:recording-modalities}
\todo{find better figure with permission to reuse and without unnecessary
modalities}

@@

\fullpagefig{figures/bci/paradigms.eps}{
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in active, reactive and passive
  paradigms, depending on their reliance on external stimulation and active
  participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{fig:bci-paradigms}
\todo{check permissions}

@@

The time-domain response ellicited immediately after intensification of a
stimulus is the visual \ac{erp}~\cite{Luck2014}.
An \ac{erp} is a waveform consisting of multiple components, of wich some are
modulated by whether the stimulus was an oddball or not.
Components are positive peaks and negative troughs in the \ac{erp} waveform.
They following a naming scheme according to their polarity (Negative or
Positive) and their latency (e.g. 3 or 300ms).
The component most prominently modulated by attention is called the P300 or P3
(positive deflection after $\pm$300ms) and oddball paradigms are therefore
frequently referred to as P300 paradigms.
However, due to the equally important contribution of other \ac{erp}
components than the P3 in decoding as we will show later, this could be
considered a bit of a misnomer and we will stick to the oddball paradigm naming.
P300 and other components, explain function (Luck)
Which components, at what amplitude and latency they are evoked varies per
person and depends on the exact details of the stimulation.

In general, earlier components, such as the \acp{vep}, are more tied to sensory
information processing and less modulated by attention, while later components
are more related to attention and cognitive processes.

\textcite{Luck2013}
\subsubsection{P1}
\subsubsection{N1}
\subsubsection{P2}
\subsubsection{N2}
\subsubsection{P3}
\subsubsection{Late negativity}

\subsection{Preprocessing and feature extraction}
referencing, filtering, epoching, baselining
\subsection{Decoding ERPs}

>>>>

\fullpagefigct{%
  \includegraphics[width=\paperwidth]{figures/bci/bci_loop.pdf}
}{%
  The \ac{bci} loop.
  The user interacts with the \ac{bci} through a \ac{bci} paradigm, in this case involving
  visual stimulation. Neural signals are recorded, and neurophysiological features
  related to the paradigm are extracted. Using machine learning, a decision can
  then be made based on these features, which can be presented back to the user.
  In a closed-loop \ac{bci}, this feedback allows the user to adapt to the \ac{bci}.
}{%
  fig:bci/loop
}

@@

The recorded brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
It would help greatly if we knew what we are looking for.
For this reason, a \ac{bci} operates using a defined \emph{paradigm}, the manner in
which interaction is performed.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
that the brain response evoked by attended stimuli is different of those
unattended,
and determine which one of a set of targets was attended.
In turn, this can be coupled to an action, like typing the letter shown on the
stimulus.
The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci-paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

@@

As a consequence, the \ac{bci} field relies heavily on the
electrical activity of the brain.
As mentioned earlier with MEG, an important concept in determining which
technology is suited for a \ac{bci} application is the \emph{invasiveness} of the technology.
As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often measure a very specific brain region and can result in better signal quality, at the cost of the risks
involved with brain surgery.

\fullpagefigct{%
  \includegraphics[width=\paperwidth]{figures/bci/recording_modalities.pdf}
}{%
	Different electrophysiology recording modalities and \\
	their respective position in a cross-section of the skull. \\
	There is a trade-off between invasiveness and  \\
	spatial resolution.
}{%
  fig:bci/recording
}
\todo{find better figure with permission to reuse and without unnecessary
modalities}

@@

\fullpagefig{%
  \input{figures/bci/paradigms.tikz.tex}
}{%
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in active, reactive and passive
  paradigms, depending on their reliance on external stimulation and active
  participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{
  fig:bci-paradigms
}
\todo{check permissions}

@@

The time-domain response elicited immediately after the intensification of a
stimulus is known as the visual \ac{erp}~\cite{Luck2014}.
An \ac{erp} is a waveform consisting of multiple components.
Some of these components are modulated by whether the stimulus was an
oddball or not.
These components appear as positive peaks and negative troughs in the \ac{erp}
waveform.
They follow a naming scheme based on polarity (Negative or Positive) and
latency (e.g., N2 or N200 for a negative component after $\pm200$ ms).
The component most prominently modulated by attention is called the P3
(positive deflection after $\pm300$ ms).
Therefore, oddball paradigms are frequently referred to as P300 paradigms.
However, due to the equally important contributions of other \ac{erp}
components in decoding, as we will show later, this could be considered a
misnomer.
Thus, we will adhere to the oddball paradigm naming.

The visual \ac{erp} components primarily include P1, N1, P2, N2, and P3.
Each component is characterized by specific latency, amplitude, and neural
origins.
These factors influence the perception of visual stimuli and attentional
mechanisms~\cite{Luck2013}.

\paragraph{P1}
The P1 component occurs approximately 100 ms after stimulus presentation.
This component reflects initial visual processing and is primarily linked to
activity in the primary visual cortex.
P1 is particularly sensitive to gaze fixation, where the gaze is directed
toward a stimulus.
When a stimulus is fixated on, the amplitude of P1 increases, reflecting the
enhanced processing of that visual input.
In contrast, P1 shows less modulation when stimuli are attended to via
attention without direct gaze.

\paragraph{N1}
The N1 component peaks around 150 to 200 ms post-stimulus.
It represents an extension of sensory processing related to both gaze fixation
and attention.
For attended stimuli, the amplitude is greater, indicating prioritization for
further cognitive processing.
This highlights a more selective form of attention allocation, demonstrating
the impact of both types of attention on visual processing.

\paragraph{P2}
The P2 component occurs between 200 to 300 ms.
It reflects higher-order processing, particularly in contexts demanding
stimulus evaluation and feature detection.
P2 is sensitive to attention.
When participants actively engage with specific features or categories of
stimuli, P2 shows increased amplitude.
Conversely, P2 may be less responsive when gaze is not directed toward the
stimulus but attention is still maintained.

\paragraph{N2}
Peaking around 200 to 350 ms, the N2 component is associated with cognitive
control processes such as conflict monitoring and inhibition.
It is particularly pronounced in tasks requiring differentiation between
competing stimuli.
N2 reflects the allocation of attentional resources required to resolve
conflicts, irrespective of whether the gaze is directly on the competing
stimuli or not.
Higher amplitudes in N2 are seen when cognitive demands increase, indicating
the influence of attention strategies.

\paragraph{P3}
The P3 component is elicited in oddball paradigms and peaks around 300 ms
post-stimulus.
It reflects attentional engagement and the processing of rare or unexpected
events.
The P3 is typically subdivided into two subcomponents: P3a and P3b.
The P3a is associated with the allocation of attention to novel or unexpected
stimuli, indicating the initial detection of a change in the environment.
It is particularly responsive to stimuli that capture attention, whether
through gaze fixation or attention.
P3a typically exhibits a frontal distribution and peaks earlier than P3b.

Conversely, the P3b relates to the evaluation of the stimulus.
It involves the updating of cognitive resources in response to task relevance.
This subcomponent is typically observed over parietal regions.
P3b is elicited when participants must process the stimulus meaningfully,
often requiring a decision or response.
Both P3a and P3b amplitudes are sensitive to the probability of occurrence
and the participant's attentional focus, showing how attentional strategies
affect cognitive processing.

\paragraph{Late Positive Potential (LPP)}
The LPP is a later component occurring beyond 400 ms.
Sustained attention and emotional processing of stimuli are reflected in the
LPP.
Typically enhanced for emotionally salient or motivationally relevant
information, LPP serves as an indicator of cognitive appraisal beyond mere
recognition tasks.
This component can be influenced by attention when the participant evaluates
the emotional significance of stimuli, even if gaze is not directed at them.

In summary, ERP components such as P1, N1, P2, N2, P3, and LPP collectively
elucidate the neural underpinnings of visual processing and attentional
mechanisms.
These components not only inform the design of an effectiv oddball BCI system
but also enhance our understanding of the cognitive processes involved in visual
attention.
This includes both gaze fixation and attention.
Understanding these components is essential for optimizing BCI protocols and
improving user interaction, particularly in contexts involving visual oddball
tasks.

\subsection{Decoding ERPs}

### 2024-09-26 chapters/gaze_independence.tex
Add ERP section

The interface can be operated in different VSA conditions as illustrated in
Figure~\ref{fig:gaze-vsa}.
\fourpanefig
  {\figpane{figures/gaze_independence/attention_overt.eps}{
    \emph{Overt \ac{vsa}}.
    Gaze and \ac{vsa} coincide on a target.}{fig:gaze-vsa-overt}}
  {\figpane{figures/gaze_independence/attention_covert.eps}{\emph{Covert
  \ac{vsa}}.
  Gaze rests on the center of the screen, while \ac{vsa} is directed towards a
  target.}{fig:gaze-vsa-covert}}
  {\figpane{figures/gaze_independence/attention_split.eps}{
    \emph{Split \ac{vsa}}. \ac{vsa} is directed towards a target, while the
    gaze rests on another.}{fig:gaze-vsa-split}}
  {\figpane{figures/gaze_independence/attention_free.eps}{\emph{Free \ac{vsa}}.
  The user is free to direct their gaze as they deem most comfortable
  All previous \ac{vsa} conditions are possible.}{fig:gaze-vsa-free}}
  {\Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.}
  {fig:gaze-vsa}

>>>>

The interface can be operated in different VSA conditions as illustrated in
Figure~\ref{fig:gaze-vsa}.
\fourpanefig{%
  \figpane{%
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.eps}
  }{%
    \emph{Overt \ac{vsa}}.
    Gaze and \ac{vsa} coincide on a target.
  }{%
    fig:gaze/vsa/overt
  }
}{%
  \figpane{%
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.eps}
  }{%
    \emph{Covert \ac{vsa}}.
    Gaze rests on the center of the screen, while \ac{vsa} is directed towards a target.
  }{%
    fig:gaze/vsa/covert
  }
}{%
  \figpane{%
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
  }{%
    \emph{Split \ac{vsa}}.
    \Ac{vsa} is directed towards a target, while the gaze rests on another.
  }{%
    fig:gaze/vsa/split
  }
}{%
  \figpane{%
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
  }{%
    \emph{Free \ac{vsa}}.
    The user is free to direct their gaze as they deem most comfortable
    All previous \ac{vsa} conditions are possible.
  }{%
    fig:gaze/vsa/free
  }
}{%
  \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
}{%
  fig:gaze-vsa
}

### 2024-09-26 chapters/wcble.tex
Add ERP section

\todo{comments liesel}
Sections~\ref{sec:wcble/literature/contrib} were published \ref{sec:wcble/methods}
by~\textcite{VanDenKerchove2024}

@@

The W\ac{cble} algorithm is presented in Algorithm~\ref{alg:wcble}.
Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using equation~\ref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{figures/wcble/figure1.pdf}{Schematic representation of the Woody Classifier Based Latency
    Estimation training phase.
    (1) The first-stage spatiotemporal binary classifier is trained on a set of
    epochs.
    (2) It is	then applied to time-shifted copies of the epochs to obtain scores
    and class probabilities	over time.
    (3) The medians of these probability distributions are assumed the class
    latencies.
    (4) The epochs are aligned to their corresponding	class latencies by
    shifting in time such that all latencies fall at the same moment.
    (5) The spatiotemporal classifier is then retrained on the aligned epochs
    for a	next iteration.
    (6) After the iterative process halts, the scores and	latencies obtained
    from the last iteration are used to train the second-stage
    classifier.}{fig:diagram-train}
The process halts after a fixed amount of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
in~Section~\ref{sec:robust-latency}.
\fullpagefig{figures/wcble/figure2.pdf}{Schematic representation of the Woody Classifier Based Latency
    Estimation test phase. (1) The
		first-stage	spatiotemporal binary classifier obtained from the training phase is
		applied to time-shifted copies of the epochs to obtain scores and class probabilities
		over time. (2) The medians of these probability distributions are assumed
		as the new class latencies. (3) The scores and class latencies are input to the
    trained second-stage classifier, which predicts the label of the epochs.}{fig:diagram-eval}
\section{Algorithms}
\todo{Pr not italic}
\todo{re-insert comments}

@@

%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{figures/wcble/simulated-sine.pdf}{}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}

>>>>

\todo{comments liesel}
Sections~\ref{sec:wcble/literature/contrib} and~\ref{sec:wcble/methods} were published
by~\textcite{VanDenKerchove2024}

@@

The W\ac{cble} algorithm is presented in Algorithm~\ref{alg:wcble}.
Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using equation~\ref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation training phase.
  (1) The first-stage spatiotemporal binary classifier is trained on a set of
  epochs.
  (2) It is	then applied to time-shifted copies of the epochs to obtain scores
  and class probabilities	over time.
  (3) The medians of these probability distributions are assumed the class
  latencies.
  (4) The epochs are aligned to their corresponding	class latencies by
  shifting in time such that all latencies fall at the same moment.
  (5) The spatiotemporal classifier is then retrained on the aligned epochs
  for a	next iteration.
  (6) After the iterative process halts, the scores and	latencies obtained
  from the last iteration are used to train the second-stage
  classifier.
}{%
  fig:diagram-train
}
The process halts after a fixed amount of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
in~Section~\ref{sec:robust-latency}.
\fullpagefig{
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation test phase. (1) The
	first-stage	spatiotemporal binary classifier obtained from the training phase is
	applied to time-shifted copies of the epochs to obtain scores and class probabilities
	over time. (2) The medians of these probability distributions are assumed
	as the new class latencies. (3) The scores and class latencies are input to the
  trained second-stage classifier, which predicts the label of the epochs.
}{%
  fig:diagram-eval
}
\section{Algorithms}
\todo{Pr not italic}
\todo{re-insert comments}

@@

%A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-gausian}, \ref{fig:wcble/sim-morlet} and
A sample of the simulated evoked data is displayed in Figures~\ref{fig:wcble/sim-sine}.
%\fullpagefig{figures/wcble/simulated-gaussian.pdf}{}{fig:wcble/sim-gaussian}
%\fullpagefig{figures/wcble/simulated-morlet.pdf}{}{fig:wcble/sim-morlet}
\fullpagefig{\includegraphics[width=\textwidth]{figures/wcble/simulated-sine.pdf}}{}{fig:wcble/sim-sine}
\todo{Properly align plots and label axes}
